---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# kerdec

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/gbasulto/kerdec.svg?branch=master)](https://travis-ci.org/gbasulto/kerdec)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/gbasulto/kerdec?branch=master&svg=true)](https://ci.appveyor.com/project/gbasulto/kerdec)
<!-- badges: end -->

`R` package to kernel deconvolution methods, including multivariate kernel deconvolution density estimation and deconvolution version of local constant polynomial regression.

## Installation

You can install the released version of kerdec from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("kerdec")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("gbasulto/kerdec")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(kerdec)
## basic example code
```

