---
title: "Kernels"
author: "Guillermo Basulto-Elias"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{kernels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r global_options, include=FALSE}
rm(list=ls()) ### To clear namespace
library(knitr)

# opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
#                echo=TRUE, warning=FALSE, message=FALSE)
opts_chunk$set(fig.width= 7, fig.height= 5,
               echo=TRUE, warning=FALSE, message=FALSE)


```

Kernel whose Fourier transform are preferred in kernel deconvolution methods due to their theoretical and numerical advantages. In fact, only those Fourier transforms are used, rather than the kernel functions themselves. Thus, we have a function that can calculates kernel and product kernels.

## Kernel functions and their Fourier transforms

The explicit expressions of kernel functions are not required in kernel deconvolution methods, but rather their Fourier transform. This package provides formulas for those Fourier transforms. We show below how they look like and then we show how do kernel functions look like, but we do use the package `fourierin` to obtain these values from their Fourier transforms.

The five kernel functions considered have boundedly supported Fourier transforms, specifically on $[-1, 1]$. We abuse on the nomenclature of the kernels, since "triangular", "tricube" and "triweight" do not correspond to the kernel functions with these names, but to the fact that their Fourier transform is proportional to these kernels. In the same manner, "flat-top" refers to a kernel whose Fourier transform is flat around zero.

```{r, cache = TRUE}
## Load packages
library(kerdec)
library(dplyr)
library(tidyr)
library(ggplot2)

expand.grid(t = seq(-1.3, 1.7, by = 0.01),
           kernel = c("sinc", "triangular", "triweight", 
                      "tricube", "flat-top")) %>%
  tbl_df() %>%
  group_by(kernel) %>%
  mutate(value = ft_kernel(t, unique(kernel))) %>%
  ggplot(aes(t, value, color = kernel)) +
  geom_line() +
  theme_bw() +
  theme(legend.position = "bottom")

```

Now we obtain the values of the kernel functions from their Fourier transforms.

```{r, cache = TRUE}
## Load package
library(kerdec)

## Write the name of all kernels available
ker <- c("sinc", "triangular", "triweight", "tricube", "flat-top")
names(ker) <- ker

## Find the actual kernel values from their Fourier transform
vals <- lapply(ker,
               function(krnl) 
                 fourierin::fourierin(function(t) ft_kernel(t, krnl), 
                                      a = -1, b = 1, c = -9, d = 9, 
                                      r = -1, s = -1, resol = 256))
## Obtain grid and extract the real part of every set of values.
x <- vals$sinc$w
vals <- lapply(vals, function(out) c(Re(out$values)))

## Generate plot
plot(x, vals[["sinc"]], type = "n", 
     ylab = "value", main = "Kernel functions")
abline(h = 0, col = "lightgray")
for(idx in 1:5){
  lines(x, vals[[idx]], col = idx + 1, lty = idx, lwd = 2)
}
legend("topright", ker, col = 2:6, lty = 1:5, lwd = 2)

```


## Univariate example

Although in the previous section we showed how to use the function that provides the Fourier transform of a particular set of kernel functions, we show a bare minimum example to illustrate how to use these function.

```{r, cache = TRUE}
## Load package 
library(fourierin)
library(kerdec)
library(dplyr)
library(ggplot2)

data_frame(t = seq(-1.3, 1.3, 0.01),
           values = ft_kernel(t, ker = "triangular")) %>%
  ggplot(aes(t, values)) +
  geom_line(color = "magenta")

## Define grid
t <- seq(-1.3, 1.3, 0.01)

## Compute values
## We obtain the same result with ker = 2
values <- ft_kernel(t, ker = "triangular")

## Print plot
plot(t, values, main = "Triangular kernel", lwd = 2, 
     type = "l", col = "magenta")
```


## Product kernels

Finally we show a minimal example of how to use the function `ft_kernel` to find the Fourier transform of a product kernel. Observe that the grid has to be introduced as a matrix with $d$ columns and not as a $d$-dimensional array. 

```{r, cache = TRUE}

## Load package 
library(fourierin)

## Define grid
t1 <- seq(-1.1, 1.1, length.out = 150)
t2 <- t1
t <- expand.grid(t1 = t1, t2 = t1)

## Compute values
## We obtain the same result with ker = 2
values <- ft_kernel(t, ker = "triangular")
values <- ft_kernel(t, ker = "flat-top")

## Convert dataframe to matrix
values <- matrix(values, length(t1))

## Print surface plot
persp(x = t1, y = t2, z = values,
      theta = 120, lwd = .2, phi = 30, shade = .8)

## Now a contour plot
contour(x = t1, y = t2, z = values, nlevels = 10)

```
